Follow-up Samples

### 2017-03-24 submission/manuscript_outline_20161024.tex

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma or SRN cases
\emph{(P-value )}. But there was a trend for increasingly less community
similarity between initial and follow up samples from adenoma to SRN to
carcinoma. Despite this, no difference was found in the relative
abundance of any OTU before and after treatment for adenoma, SRN, or
carcinoma groups \emph{(P-value )}. Using Random Forest models to assess
whether changes in follow up samples were towards a normal community
showed that only those with carcinoma had a significant decrease in
positive probability \emph{(P-value)}; providing further evidence that
treatment has the greatest effect in those with carcinoma. The adenoma
model used a total of \emph{model n} OTUs, the SRN model used a total of
\emph{model n} OTUs, and the carcinoma model used a total of \emph{model
n} OTUs. A total of \emph{common n} OTUs were common to both models with
many classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Anaerostipes}, \emph{Blautia}, and
\emph{Dorea}). Both chemotherapy and radiation did not provide any
additional changes to the bacterial community in those treated for
carcinoma \emph{(Pvalue )}

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether those with adenoma \emph{(n = )}, SRN \emph{(n = )}, or
carcinoma \emph{(n = )} had any broad differences between their initial
and follow up samples. We found that those with carcinoma had a more
dissimilar bacterial community between their initial and follow up
sample than those with adenoma \emph{(P-value 0.001)} {[}Figure 1A{]}.
Although no significant differences were observed between SRN and
carcinoma there was an increase in the dissimilarity of the initial and
follow up samples from adenoma \emph{(mean and SD)} to SRN \emph{(mean
and SD)} to carcinoma \emph{(mean and SD)} {[}Figure 1A{]}. The
bacterial community structure before and after surgery was visualized
using NMDS for adenoma {[}Figure 1B{]} \emph{(PERMANOVA 0.05)}, SRN
{[}FIgure 1C{]} \emph{(PERMANOVA 0.05)}, and carcinoma {[}Figure 1C{]}
\emph{(PERMANOVA 0.05)}. Interestingly, when initial and follow up
samples were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
\emph{(PERMANOVA 0.05)}. There was no difference between initial and
follow up samples when investigating alpha diversity metrics for
adenoma, SRN, or carcinoma for any metric tested {[}Table S1{]}.
Additionally, there was also no difference in the relative abundance of
any OTU between initial and follow up samples for adenoma, SRN, or
carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model}} The range of model AUC's from 100 runs of
20 repeated 10 fold cross-validation was \emph{from to this} with the
AUC of the model used being \emph{AUC value} {[}Figure S2A{]}. There was
a total of \emph{blank} OTUs in this model with the vast majority
classifying to bacteria typically thought of as commensal {[}Figure
S3A{]}. There was a significant difference between the actual and model
predicted group stratification calls \emph{(P-value)}. There was no
significant decrease in the positive probability of adenoma between
initial and follow up samples \emph{(P\_value)} {[}Figure 2A{]}.

\textbf{\emph{SRN Model}} The range of model AUC's from 100 runs of 20
repeated 10 fold cross-validation was \emph{from to this} with the AUC
of the model used being \emph{AUC value} {[}Figure S2B{]}. There was a
total of \emph{blank} OTUs in the SRN model {[}Figure S3B{]}. Similar to
the adenoma model the vast majority of OTUs classified to bacteria
typically thought of as commensal. Also similar to the adenoma model
there was a significant difference between the actual and model
predicted group stratification calls \emph{(P-value)}. There was no
significant decrease in the positive probability of SRN between initial
and follow up samples \emph{(P\_value)} {[}Figure 2B{]}.

\textbf{\emph{Carcinoma Model}} The range of model AUC's from 100 runs
of 20 repeated 10 fold cross-validation was \emph{from to this} with the
AUC of the model used being \emph{AUC value} {[}Figure S2C{]}.
Interestingly, the AUCs improved from adenoma to SRN to carcinoma
{[}Figure 2{]}. There was a total of \emph{blank} OTUs in the carinoma
model {[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
majority of OTUs classified to bacteria typically thought of as
commensal but OTUs that also classified to \emph{Fusobacterium},
\emph{Porphyromonas}, and \emph{Parvimonas} appear to be important for
carcinoma classification {[}Figure S3C{]}. Also similar to the adenoma
and SRN models there was a significant difference between the actual and
model predicted group stratification calls \emph{(P-value)}. There was a
significant decrease in the positive probability of carcinoma between
initial and follow up samples \emph{(P\_value)} {[}Figure 2C{]};
suggesting that the carcinoma samples changed towards normal after
treatment, unlike either adenoma or SRN. The one indivdiual still
positive for carcinoma after treatment had an increase in carcinoma
positive proability on follow up {[}Figure 2C{]}.

\textbf{\emph{Adenoma, SRN, and Carcinoma Common OTUs}} We next wanted
to know what predictors within the adenoma, SRN, and carcinoma models
were similar to each other. The main purpose was to identify which OTUs
could be important at all three stages of disease. When we compared the
three different models with each other there were a total of
\emph{number} common OTUs. Some of the most common taxonomic
identifications belonged to \emph{one}, \emph{two}, \emph{three},
\emph{four}, \emph{five}, \emph{six}, and \emph{seven}. These along with
the vast majority of the OTUs that were common between these models had
classifications to bacteria typically thought of as commensal {[}Table
S2{]}.

\textbf{\emph{Treatment Affects on Community}} After observing these
these changes from treatment we assessed whether chemotherapy or
radiation, in the carcinoma group, and surgery, in the adenoma group,
impacted the observed results. In the carcinoma group neither
chemotherapy or radiation provided any additive change from initial
sample over surgery alone {[}Table S3{]}. For the adenoma group there
was no difference between those that received surgery and those that did
not {[}Table S4{]}. For surgery adenoma and SRN were combined due to the
low number of surgery occurances. There was no difference in the
proportion of those receiving surgery between the adenoma and SRN groups
\emph{(P-value)}. This data suggests that microbiome changes observed in
the carcinoma group were mostly a result from surgery and not from
chemotherapy or radiation.

@@

Many of the common OTUs that we identified taxnomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}, and \emph{Anaerostipes}){[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}14{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}15{]}. High protein consumption in general has been linked with an
increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}17{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma or SRN versus carcinoma \emph{(uncorrected P-value 0.05)}, with
time passed being less for adenoma \emph{( days)} and SRN \emph{( days)}
than carcinoma \emph{( days)}. These results would indicate that some of
the differences observed between the carcinoma and adenoma groups could
be due to differences in collection time. Specifically, it could
confound the observation that carcinomas changed more than adenomas
{[}Figure 1A \& 1C{]}. However, there are two reasons that this may not
be the case. First, the SRN group did not have a significant difference
and the collection time of their follow up sample was less than the
adenoma group. Second, this confounding would not affect the
observations where models were used since they were built using a
different cohort {[}Figures 2 \& S2-S3{]}.

>>>>

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma (P-value
\textless{} 0.05) but not carcinoma versus SRN cases (P-value
\textgreater{} 0.05). But there was a trend for increasingly less
community similarity between initial and follow up samples from adenoma
to SRN to carcinoma. Despite this, no difference was found in the
relative abundance of any OTU before and after treatment for adenoma,
SRN, or carcinoma groups (P-value \textgreater{} 0.05). Using Random
Forest models to assess whether changes in follow up samples were
towards a normal community showed that only those with carcinoma had a
significant decrease in positive probability (P-value \textless{} 0.05);
providing further evidence that treatment has the greatest effect in
those with carcinoma. The adenoma model used a total of 62 OTUs, the SRN
model used a total of 61 OTUs, and the carcinoma model used a total of
59 OTUs. A total of 26 OTUs were common to both models with many
classifying to commensal bacteria (e.g. \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Anaerostipes}, \emph{Blautia}, and
\emph{Dorea}). Both chemotherapy and radiation did not provide any
additional changes to the bacterial community in those treated for
carcinoma (Pvalue \textgreater{} 0.05).

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether those with adenoma (n = 22), SRN (n = 19), or carcinoma
(n = 26) had any broad differences between their initial and follow up
samples. We found that those with carcinoma had a more dissimilar
bacterial community between their initial and follow up sample than
those with adenoma (P-value \textless{} 0.001) {[}Figure 1A{]}. Although
no significant differences were observed between SRN and carcinoma there
was an increase in the dissimilarity of the initial and follow up
samples from adenoma (0.55 ± 0.21 (mean ± SD)) to SRN (0.65 ± 0.25) to
carcinoma (0.78 ± 0.15) {[}Figure 1A{]}. The bacterial community
structure before and after surgery was visualized using NMDS for adenoma
{[}Figure 1B{]} (PERMANOVA \textgreater{} 0.05), SRN {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
(PERMANOVA \textless{} 0.05). Interestingly, when initial and follow up
samples were compared, regardless of whether the lesions were adenoma or
carcinoma, there was no significant overall difference in beta diversity
(PERMANOVA \textgreater{} 0.05). There was no difference between initial
and follow up samples when investigating alpha diversity metrics for
adenoma, SRN, or carcinoma for any metric tested {[}Table S1{]}.
Additionally, there was also no difference in the relative abundance of
any OTU between initial and follow up samples for adenoma, SRN, or
carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model}} The range of model AUC's from 100 runs of
20 repeated 10 fold cross-validation was 0.62 - 0.72 with the AUC of the
model used for classification havin an AUC of 0.65 {[}Figure S2A{]}.
There was a total of 62 OTUs in this model with the vast majority
classifying to bacteria typically thought of as commensal {[}Figure
S3A{]}. There was a significant difference between the actual and model
predicted group stratification calls (P-value \textless{} 0.05). There
was no significant decrease in the positive probability of adenoma
between initial and follow up samples (P-value \textgreater{} 0.05)
{[}Figure 2A{]}.

\textbf{\emph{SRN Model}} The range of model AUC's from 100 runs of 20
repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
There was a total of 61 OTUs in the SRN model {[}Figure S3B{]}. Similar
to the adenoma model the vast majority of OTUs classified to bacteria
typically thought of as commensal. Also similar to the adenoma model
there was a significant difference between the actual and model
predicted group stratification calls (P-value \textless{} 0.05). There
was no significant decrease in the positive probability of SRN between
initial and follow up samples (P-value \textgreater{} 0.05) {[}Figure
2B{]}.

\textbf{\emph{Carcinoma Model}} The range of model AUC's from 100 runs
of 20 repeated 10 fold cross-validation was 0.84 - 0.9 with the AUC of
the model used for classification being 0.88 {[}Figure S2C{]}.
Interestingly, the AUCs improved from adenoma to SRN to carcinoma
{[}Figure 2{]}. There was a total of 59 OTUs in the carinoma model
{[}Figure S3C{]}. Similar to the adenoma and SRN models the vast
majority of OTUs classified to bacteria typically thought of as
commensal but OTUs that also classified to \emph{Fusobacterium},
\emph{Porphyromonas}, and \emph{Parvimonas} appear to be important for
carcinoma classification {[}Figure S3C{]}. Also similar to the adenoma
and SRN models there was a significant difference between the actual and
model predicted group stratification calls (P-value \textless{} 0.05).
There was a significant decrease in the positive probability of
carcinoma between initial and follow up samples (P-value \textless{}
0.05) {[}Figure 2C{]}; suggesting that the carcinoma samples changed
towards normal after treatment, unlike either adenoma or SRN. The one
indivdiual still positive for carcinoma after treatment had an increase
in carcinoma positive proability on follow up {[}Figure 2C{]}.

\textbf{\emph{Adenoma, SRN, and Carcinoma Common OTUs}} We next wanted
to know what predictors within the adenoma, SRN, and carcinoma models
were similar to each other. The main purpose was to identify which OTUs
could be important at all three stages of disease. When we compared the
three different models with each other there were a total of 26 common
OTUs. Some of the most common taxonomic identifications belonged to
\emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
\emph{Lachnospiraceae}, and \emph{Dorea}. These along with the vast
majority of the OTUs that were common between these models had
classifications to bacteria typically thought of as commensal {[}Table
S2{]}.

\textbf{\emph{Treatment Affects on Community}} After observing these
these changes from treatment we assessed whether chemotherapy or
radiation, in the carcinoma group, and surgery, in the adenoma group,
impacted the observed results. In the carcinoma group neither
chemotherapy or radiation provided any additive change from initial
sample over surgery alone (P-value \textgreater{} 0.05) {[}Table S3{]}.
For the adenoma group there was a single difference in observed OTUs
(sobs) between those that received surgery and those that did not
(P-value \textless{} 0.05) {[}Table S4{]}. For the surgery comparison,
adenoma and SRN were combined due to the low number of surgery
occurances in these two groups. There was no difference in the
proportion of those receiving surgery between the adenoma and SRN groups
(P-value \textgreater{} 0.05). This data suggests that microbiome
changes observed in the carcinoma group were mostly a result from
surgery and not from chemotherapy or radiation.

@@

Many of the common OTUs that we identified taxnomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}, and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}14{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}15{]}. High protein consumption in general has been linked with an
increased CRC risk {[}16{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}17{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression.

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma or SRN versus carcinoma (P-value \textless{} 0.05), with time
passed being less for adenoma (255 ± 42 days) and SRN (250 ± 41) than
carcinoma (351 ± 102). These results would indicate that some of the
differences observed between the carcinoma and adenoma groups could be
due to differences in collection time. Specifically, it could confound
the observation that carcinomas changed more than adenomas {[}Figure
1{]}. However, there are two reasons that this may not be the case.
First, the SRN group did not have a significant difference and the
collection time of their follow up sample was less than the adenoma
group. Second, this confounding would not affect the observations where
models were used since they were built using a different cohort
{[}Figures 2 \& S2-S3{]}.

### 2017-03-24 submission/supplemental_outline_20161024.Rmd

>>>>

**Table S4: Summary of Adenoma based on Surgery Received**

```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}

adn_surg_summary <- read.csv('../data/process/tables/adn_combined_probs_surgery_pvalue_summary.csv', 
                              header = T, stringsAsFactors = F) %>% 
  filter(X != "prop_surg_adn_srn") %>% 
  mutate(variable = c("sobs", "Shannon", "Evenness", "Thetayc", "Adenoma Only\nPositive Probability", "SRN Only\nPositive Probablity"))

st4 <- cbind(adn_surg_summary$variable, format(adn_surg_summary$surg_mean, digits = 1, nsmall = 1), format(adn_surg_summary$no_surg_mean, digits = 1, nsmall = 1), format(adn_surg_summary$surg_sd, digits = 1, nsmall = 1), format(adn_surg_summary$no_surg_sd, digits = 1, nsmall = 1), format(adn_surg_summary$pvalue, scientific = TRUE, digits = 2), format(adn_surg_summary$bh, scientific = TRUE, digits = 2))

colnames(st4) <- c("Variable", "Surgery Mean", "No Surgery Mean", "Surgery SD", "No Surgery SD", "P-value", "BH Corrected")

kable(st4, format = "markdown", align = 'c', padding = 2)
